A weakly structured stem for human origins in Africa

### 2022-12-23 paper.tex
Revisions to address Reviewer 5

Decades of study of human genome variation have suggested a predominantly
tree-like model of recent population divergence from a single ancestral
population in Africa. It has been difficult to reconcile this finding with the
fossil and archaeological records of human occupation across the vast African
continent. For example, fossils such as those from the sites of Jebel Irhoud,
Morocco \citep{Hublin2017-cq}, Herto, Ethiopia \citep{White2003-bk} and Klasies
River, South Africa \citep{Deacon1995-rx} demonstrate that derived \emph{Homo
sapiens} anatomical features were intermittently present across the continent
300-100ka. Archaeological sites from the Middle Stone Age, of which some have
been associated with \emph{Homo sapiens}, are also widely distributed across
Africa. It is unclear whether these fossils and archaeological sites represent
populations which contributed to contemporary \emph{Homo sapiens} as population
precedents or were local ``dead-ends''. Recently, attempts to reconcile genetic
and paleoanthropological data include proposals for a Pan-African origin of
\emph{Homo sapiens} by which populations in many regions of the continent
contributed to the formation of \emph{Homo sapiens} beginning at least 300ka
\citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.

@@

Multiple studies have shown a correspondence between phenotypic
differentiation, usually assessed with measurements of the cranium, and genetic
differentiation among human populations and between humans and Neanderthals
\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb}
(see also SI Section~\ref{sec:morpho_diff}).
This correspondence allows predictions of our model to be related to the fossil record.
The fossil record of Africa is sparse during the time period of the stems, but of the
available fossils, some are very similar in morphology to contemporary humans
(e.g., from Omo Kibish, Ethiopia \citep{Day1969-rh,Vidal2022-qe}),
others are similar in some morphological features but not others
(e.g., from Jebel Irhoud, Morocco \citep{Hublin2017-cq,Richter2017-zu}),
and others are very different in morphology
(e.g., from Dinaledi, South Africa \citep{Berger2015-bq,Dirks2017-uk}).
If, as our model predicts, the
genetic differences between the stems were comparable to those among
contemporary human populations, the most morphologically divergent fossils are unlikely to
represent branches that contributed appreciably to contemporary human
ancestries.

>>>>

Decades of study of human genome variation have suggested a predominantly
tree-like model of recent population divergence from a single ancestral
population in Africa. It has been difficult to reconcile this finding with the
fossil and archaeological records of human occupation across the vast African
continent. For example, fossils such as those from the sites of Jebel Irhoud,
Morocco \citep{Hublin2017-cq}, Herto, Ethiopia \citep{White2003-bk} and Klasies
River, South Africa \citep{Deacon1995-rx} demonstrate that derived \emph{Homo
sapiens} anatomical features appeared across the continent
300-100ka. Archaeological sites from the Middle Stone Age, of which some have
been associated with \emph{Homo sapiens}, are also widely distributed across
Africa. It is unclear whether these fossils and archaeological sites represent
populations which contributed to contemporary \emph{Homo sapiens} as population
precedents or were local ``dead-ends''. Recently, attempts to reconcile genetic
and paleoanthropological data include proposals for a Pan-African origin of
\emph{Homo sapiens} by which populations in many regions of the continent
contributed to the formation of \emph{Homo sapiens} beginning at least 300ka
\citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.

@@

Multiple studies have shown a correspondence between phenotypic differentiation, 
usually assessed with measurements of the cranium, and genetic differentiation 
among human populations and between humans and Neanderthals 
\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also SI 
Section~\ref{sec:morpho_diff}). This correspondence potentially allows predictions 
of our model to be related to the fossil record. The fossil record of Africa is 
sparse during the time period of the stems, but of the fossils that date to this 
time period, some are fairly similar in morphology to contemporary humans (e.g., 
Omo 1 from Omo Kibish, Ethiopia \citep{Day1969-rh,Vidal2022-qe}), others are 
similar in some morphological features to contemporary humans (e.g., Irhoud 1 
from Jebel Irhoud, Morocco \citep{Hublin2017-cq,Richter2017-zu}), and others are 
different enough in morphology to have been assigned to species other than 
\emph{Homo sapiens} (e.g., DH1 from Dinaledi, South Africa 
\citep{Berger2015-bq,Dirks2017-uk}). If, as our model predicts, the genetic 
differences between the stems were comparable to those among contemporary human 
populations, the most morphologically divergent fossils are unlikely to represent 
branches that contributed appreciably to contemporary human ancestries.
